---
title: "Using Plotly to Create Interactive Graphs"
subtitle: "Example"
author:  "Gerard Torrats-Espinosa"
date: "Fall 2020"
output:
  html_document:
    number_sections: false
    toc: false
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
---


<style type="text/css">

h1.title { /* Title */
    font-size: 24px;
    color: DarkBlue;
    font-weight: bold;
    font-style: normal;
}
h3.subtitle { /* Subtitle */
    font-size: 20px;
    color: DarkBlue;
    font-weight: normal;
    font-style: normal;
}
h4.author { /* Author */
    font-size: 16px;
    color: Gray;
    font-weight: normal;
    font-style: normal;
}
h4.date { /* Date */
    font-size: 14px;
    color: Gray;
    font-weight: normal;
    font-style: italic;
}


h1 { /* Header 1 */
    font-size: 20px;
    color: DarkBlue;
    font-weight: bold; <!-- this could be:  font-weight: bold -->
    font-style: normal;
}
h2 { /* Header 2 */
    font-size: 18px;
    color: DarkBlue;
    font-weight: bold;
    font-style: normal;
}
h3 { /* Header 3 */
    font-size: 16px;
    color: DarkBlue;
    font-weight: normal;
    font-style: normal;
}
h4 { /* Header 4 */
    font-size: 14px;
    color: DarkBlue;
    font-weight: normal;
    font-style: normal;
}
h5 { /* Header 5 */
    font-size: 12px;
    color: Black;
    font-weight: normal;
    font-style: normal;
}


body{ /* Normal  */
      font-size: 12px;
      font-family: "Verdana", Times, serif;
  }
td {  /* Table  */
  font-size: 8px;
}


code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Example of interactive plot.



## Upload Packages

```{r}
library(plotly)

```


## First Plot

```{r}
p1 <- plot_ly(economics, x = ~date, y = ~unemploy / pop)
p1
```

## Second Plot

```{r}
p2 <- plot_ly(economics, x = ~date, y = ~unemploy)
p2
```